The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-14 script/en/chapters/numbertheory.tex
fixed \footnotemark in section heading introduced in 2694
incorporated user feedback, removed some \newpage

% --------------------------------------------------------------------------
\vskip +40pt
\subsection{Security parameters because of new algorithms}
\label{chptSecurityParam}

%\vskip +15 pt
%\paragraph{Factorisation algorithms\footnotemark}
\paragraph[Factorisation algorithms]
          {Factorisation algorithms\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can gather more
    experience with current factorization algorithms
    (see learning unit 5.1-5.5, pages 1-15/15).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\mbox{}

The complexity is basically determined by the length $l$ of the module $n$.
Higher values for this major parameter are oriented at the possibilities of the
current algorithms for factorization:

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & & {\bf Factored on} & {\bf Factored by} \\
\hline
	RSA-155	&	155 & 512 & & August, 1999 & Herman te Riele et al. \\
	C158 & 		158 & 523 & & January, 2002 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & & April, 2003 & Jens Franke et al. \\
	RSA576 & 	174 & 576 & & December, 2003 & Jens Franke et al. \\
	C176 & 		176 & 583 & & May, 2005 & Kazumaro Aoki et al. \\
	RSA640\footnotemark & 	193 & 600 & & November, 2005 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & & May, 2005 & Jens Franke et al. \\
	C307 & 		307 & 1017 & & May, 2007 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of July 2009)}    % Eyecatcher_neue-Mersenne 
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.\\
RSA Labs offers its challenges since the beginning of the 1990th.
The next bigger challenge awarded with 30,000 US dollars is RSA-704.\\
See:
\href{http://www.heise.de/newsticker/meldung/print/65957}
{\texttt{http://www.heise.de/newsticker/meldung/print/65957}}
}

@@

The task of the researchers from Bonn was not initiated by a challenge, but
they wanted to find the last prime factors of the integer $2^{953} - 1$
(see ``Wanted List'' of the Cunningham Project\index{%
Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}}).

@@

The task of the researchers was not initiated by a challenge, but
they wanted to find the last prime factors of the Mersenne number $2^{1039}+1$
(see ``Wanted List'' of the Cunningham Project\index{%
Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}\\
Cunningham table: \href{http://homes.cerias.purdue.edu/~ssw/cun/pmain1206}
                {\texttt{http://homes.cerias.purdue.edu/\~{}ssw/cun/pmain1206}}\\
The numbers in the Cunningham table have the following syntax:\\
``(2,n)-'' means $2^{n}-1$;~~~
``(2,n)+'' means $2^{n}+1$.\\
To describe the magnitude one writes $p<n>$ or $c<n>$: 
``n'' is the number of decimal digits and ``p'' and ``c'' tell,
whether the number is prime or composite.\\
$2^{1039}-1 = p7 * c307 = p7 * p80 * p227$ \\
It is explained more precisely at the CUN page:\\
``2651+ means $2^{651} + 1$ and the size (c209 means 209 decimal digits)
of the
number which was factored.  Then come the new factor(s), the discoverer and
the method used.  Recently, only the multiple polynomial quadratic sieve
(ppmpqs), the elliptic curve method (ecm) and the number field sieve (nfs)
have been used.  `hmpqs' stands for hypercube multiple polynomial quadratic
sieve.  Under `new factors', `p90' means a 90-digit prime and `c201' is a
201-digit composite number.''.
}).\\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_12}{}
\section{The RSA procedure with actual numbers\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Additional material: Minh Van Nguyen: "Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography", 2009.
    Didactically very clear article about some basics in NT and in Sage usage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsaconcrete}\index{RSA}
Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

>>>>

% --------------------------------------------------------------------------
\vskip +40pt
\subsection{Security parameters because of new algorithms}
\label{chptSecurityParam}
\nopagebreak
%\vskip +15 pt
%\paragraph{Factorisation algorithms\footnotemark}
\paragraph[Factorisation algorithms]
          {Factorisation algorithms\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can gather more
    experience with current factorization algorithms
    (see learning unit 5.1-5.5, pages 1-15/15).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\nopagebreak
The complexity is basically determined by the length $l$ of the module $n$.
Higher values for this major parameter are oriented at the possibilities of the
current algorithms for factorization:

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & & {\bf Factored on} & {\bf Factored by} \\
\hline
	C307 & 		307 & 1017 & & May, 2007 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & & May, 2005 & Jens Franke et al. \\
	RSA640\footnotemark & 	193 & 600 & & November, 2005 & Jens Franke et al. \\
	C176 & 		176 & 583 & & May, 2005 & Kazumaro Aoki et al. \\
	RSA576 & 	174 & 576 & & December, 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & & April, 2003 & Jens Franke et al. \\
	C158 & 		158 & 523 & & January, 2002 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & & August, 1999 & Herman te Riele et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of July 2009)}    % Eyecatcher_neue-Mersenne 
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.\\
RSA Labs offers its challenges since the beginning of the 1990th.
The next bigger challenge awarded with 30,000 US dollars is RSA-704.\\
See:
\href{http://www.heise.de/newsticker/meldung/print/65957}
{\texttt{http://www.heise.de/newsticker/meldung/print/65957}}
}

@@

The task of the researchers from Bonn was not initiated by a challenge, but
they wanted to find the last prime factors of the integer $2^{953} - 1$
(see ``Wanted List'' of the Cunningham Project\index{Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}}).

@@

The task of the researchers was not initiated by a challenge, but
they wanted to find the last prime factors of the Mersenne number $2^{1039}+1$
(see ``Wanted List'' of the Cunningham Project\index{Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}\\
Cunningham table: \href{http://homes.cerias.purdue.edu/~ssw/cun/pmain1206}
                {\texttt{http://homes.cerias.purdue.edu/\~{}ssw/cun/pmain1206}}\\
The numbers in the Cunningham table have the following syntax:\\
``(2,n)-'' means $2^{n}-1$;~~~
``(2,n)+'' means $2^{n}+1$.\\
To describe the magnitude one writes $p<n>$ or $c<n>$: 
``n'' is the number of decimal digits and ``p'' and ``c'' tell,
whether the number is prime or composite.\\
$2^{1039}-1 = p7 * c307 = p7 * p80 * p227$ \\
It is explained more precisely at the CUN page:\\
``2651+ means $2^{651} + 1$ and the size (c209 means 209 decimal digits)
of the
number which was factored.  Then come the new factor(s), the discoverer and
the method used.  Recently, only the multiple polynomial quadratic sieve
(ppmpqs), the elliptic curve method (ecm) and the number field sieve (nfs)
have been used.  `hmpqs' stands for hypercube multiple polynomial quadratic
sieve.  Under `new factors', `p90' means a 90-digit prime and `c201' is a
201-digit composite number.''.
}).\\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_12}{}
\section[The RSA procedure with actual numbers]{The RSA procedure with actual numbers\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Additional material: Minh Van Nguyen: "Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography", 2009.
    Didactically very clear article about some basics in NT and in Sage usage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsaconcrete}\index{RSA}
Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

### 2009-07-14 script/en/chapters/paper_and_pencil.tex
incorporated user feedback, removed some \newpage

\item {\bf Nihilist Substitution}\footnote{An animation of this Nihilist method
   can be find in CrypTool at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix and replace 
   every letter of the message with the two corresponding digits. The 
   resulting numbers are written into a grille. For this purpose, a 
   keyword is chosen and placed above the columns of the grille. Its 
   letters are substituted by numbers, too. The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
%------------------------------------------------------------------------------
\subsection{Homophonic substitution ciphers}

@@

\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
%------------------------------------------------------------------------------
\section{Further methods}
\label{Further-PaP-methods}

>>>>

\item {\bf Nihilist Substitution}\footnote{An animation of this Nihilist method
   can be find in CrypTool at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix to assign each letter the number built 
   from row and column number.  
   A keyword is chosen and placed above the columns of a second matrix (grille). 
   The plaintext is written row by row into the grille.
   The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

%------------------------------------------------------------------------------
\subsection{Homophonic substitution ciphers}

@@

%------------------------------------------------------------------------------
\section{Further methods}
\label{Further-PaP-methods}

### 2009-07-14 script/en/chapters/primes.tex
incorporated user feedback, removed some \newpage

% be_2005_UPDATEN_if-new-mersenne-prime-appears
For the next prize of 100,000 USD offered by EFF for a proven prime with more
than 10 million decimal digits, Edson Smith qualified, who found within the
GIMPS project $ 2^43,112,609-1 $.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

>>>>

% be_2005_UPDATEN_if-new-mersenne-prime-appears
For the next prize of 100,000 USD offered by EFF for a proven prime with more
than 10 million decimal digits, Edson Smith qualified, who found $ 2^{43,112,609}-1 $ within the
GIMPS project.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

### 2009-07-14 script/en/script-en.tex
moved "Using Sage" into appendix and "GNU License" before list of figures; added pdf bookmark for Contents
incorporated user feedback, removed some \newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Hack, damit Aufruf von \tableofcontents auch gleich \addcontentsline
% enthält. Damit steht im Inhaltsverzeichnis das Inhaltsverzeichnis
% mit korrekter Seitenzahl und Hyperlink.
% \makeatletter
% \renewcommand\tableofcontents{%
%     \section*{\contentsname
%         \@mkboth{%
%            \MakeUppercase\contentsname}{\MakeUppercase\contentsname}}%
%     \addcontentsline{toc}{section}{\contentsname}
%     \@starttoc{toc}%
%     }
% \makeatother

@@

\begin{appendix}
  \newpage
  \input{chapters/menus.tex}
  \input{chapters/authors.tex}
  \input{chapters/gnu-fdl-1.3.tex}
  \input{chapters/what-is-sage.tex}
\end{appendix}

>>>>

@@

\begin{appendix}
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{appendix-start}{}\label{s:appendix-start}
\chapter{Appendix}
    \begin{itemize}
      \item[1] \hyperlink{appendix-menutree}{CrypTool Menu Tree}
      \item[2] \hyperlink{appendix-authors}{Authors of the CrypTool Script}
      \item[3] \hyperlink{appendix-authors}{Bibliography of Movies and
                     Fictional Literature with Relation to Cryptography,
                     Books for Kids with Collections of Simple Ciphers}
      \item[4] \hyperlink{appendix-Learn-NT}
                     {Learning Tool for Elementary Number Theory}
      \item[5] \hyperlink{appendix-using-sage}
                     {Using Sage with this Script}
    \end{itemize}
  \newpage
  \input{chapters/menus.tex}
  \input{chapters/authors.tex}
  \input{chapters/what-is-sage.tex}
\end{appendix}

\input{chapters/gnu-fdl-1.3.tex}
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\clearpage
\listoffigures
\addcontentsline{toc}{chapter}{\listfigurename}

\clearpage
\listoftables
\addcontentsline{toc}{chapter}{\listtablename}
